---
title: 'Homework #2'
author: "Max McClure"
date: "2025-11-08"
output: html_document
---

# loading in libraries and settings for knit
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/users/m/r/mrmcclur/projects/eco_genomics_2025/transcriptomics/mydata")

## Import the libraries that we're likely to need in this session

library(WGCNA)
library(DESeq2)
library(tidyverse)
library(CorLevelPlot)
library(gridExtra) 
library(Rmisc)
```

## initial code chunk: read in counts table, run DEseq2, variance stabilization, read in trait data, and run WGCNA
```{r}

countsTable <- read.csv("counts_table.csv", header=TRUE, row.names=1)
dim(countsTable)

countsTableRound <- round(countsTable) # bc DESeq2 doesn't like decimals (and Salmon outputs data with decimals)
# Corrects the column names to match between the metadata table and the counts matrix.
colnames(countsTableRound) <- substr(colnames(countsTableRound), start=1, stop = 4)

#import the sample description table
conds <- read.delim("metadata.txt", header=TRUE, stringsAsFactors = TRUE, row.names=1)


#### Create a DESeq object and define the experimental design here with the tilda

dds <- DESeqDataSetFromMatrix(countData = countsTableRound, colData=conds, 
                              design= ~ generation + line)

dim(dds)

# Filter out genes with too few reads - remove all genes with counts < 15 in more than 75% of 24 samples, so 18)
## suggested by WGCNA on RNAseq FAQ

dds <- dds[rowSums(counts(dds) >= 15) >= 18,]
nrow(dds) 

dds$line <- relevel(dds$line, ref = "control")
dds$generation <- relevel(dds$generation, ref = "G1")

dds$group <- factor(paste0(dds$line, dds$generation))
design(dds) <- ~ group
dds <- DESeq(dds)

# transform the data using variance stabilization
vsd <- vst(dds, blind=FALSE)

ULT_trait <- read.csv("WGCNA_TraitData.csv", 
                      header = TRUE, 
                      stringsAsFactors = FALSE,
                      row.names = 1)
                      
# Make sure ULT_trait is a data frame
ULT_trait <- as.data.frame(ULT_trait)


#make sure you have the varience stabilized dds object 

# transform normalized counts
norm.counts <- assay(vsd) %>% 
  t()

#filter to only look at the samples that we have data for 
# Filter only samples present in norm.counts
ULT_trait <- ULT_trait[rownames(ULT_trait) %in% rownames(norm.counts), , drop = FALSE]

# calculate variance for each gene
gene_vars <- apply(norm.counts, 2, var)

# keep top 50% most variable genes (adjust proportion as needed)
n_keep <- floor(length(gene_vars) * 0.5) #
high_var_genes <- names(sort(gene_vars, decreasing = TRUE))[1:n_keep]

# subset matrix
norm.counts <- norm.counts[, high_var_genes]

load("bwnet_results_WGCNA.RData") #to load the bwnet file back in 

```

## visualize powers
```{r}
# Choose a set of soft-thresholding powers
power <- c(c(1:10), seq(from = 12, to = 50, by = 2)) #calculate the networks by a variety of powers
sft <- pickSoftThreshold(norm.counts, 
                         powerVector = power,
                         networkType = "signed",
                         verbose = 5)

sft.data <-sft$fitIndices
#visualization to pick power

a1 <- ggplot(sft.data, aes(Power, SFT.R.sq, label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  geom_hline(yintercept = 0.8, color = 'red') +
  labs(x = 'Power', y = 'Scale free topology model fit, signed R^2') +
  theme_classic()


a2 <- ggplot(sft.data, aes(Power, mean.k., label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  labs(x = 'Power', y = 'Mean Connectivity') +
  theme_classic()


grid.arrange(a1, a2, ncol = 2)

```

## load WGCNA results (ran in class) and revisualize the dendrogram
```{r}
norm.counts[] <- sapply(norm.counts, as.numeric)
dim(norm.counts) # check dimensions: 
soft_power <- 19
temp_cor <- cor
cor <- WGCNA::cor

load("bwnet_results_WGCNA.RData") #to load the bwnet file back in 

module_eigengenes <- bwnet$MEs

# Print out a preview
head(module_eigengenes)

# get number of genes for each module
table(bwnet$colors)

# Plot the dendrogram and the module colors before and after merging underneath
plotDendroAndColors(bwnet$dendrograms[[1]], cbind(bwnet$unmergedColors, bwnet$colors),
                    c("unmerged", "merged"),
                    dendroLabels = FALSE,
                    addGuide = TRUE,
                    hang= 0.03,
                    guideHang = 0.05)
```

## visualize significant eigengenes
  - MElightcyan is significant
```{r}
nSamples <- nrow(norm.counts) #
nGenes <- ncol(norm.counts) #
# correlation between module eigengenes and trait (ULT)
module.trait.corr <- cor(module_eigengenes, ULT_trait, use = 'p')
module.trait.corr.pvals <- corPvalueStudent(module.trait.corr, nSamples)

# merge eigengenes with trait data
heatmap.data <- merge(module_eigengenes, ULT_trait, by = "row.names") %>%
  column_to_rownames(var = "Row.names")

# automatically detect eigengene and trait columns
eigengene_cols <- grep("^ME", colnames(heatmap.data), value = TRUE)
trait_cols <- colnames(ULT_trait)

# plot correlations
CorLevelPlot(heatmap.data,
             x = trait_cols,
             y = eigengene_cols,
             col = c("blue1", "skyblue", "white", "pink", "red"))
```

## plot our particular eigengene across all generations
```{r}
module.gene.mapping <- as.data.frame(bwnet$colors) # assigns module membership to each gene
module.gene.mapping %>% 
  filter(`bwnet$colors` == 'MElightcyan') %>% 
  rownames()

groups <- conds[,c(1,2)]
module_eigengene.metadata <- merge(groups, heatmap.data, by = 'row.names')

library(Rmisc)
#Create a summary data frame of a particular module eigengene information
MElightcyan_summary <- summarySE(module_eigengene.metadata, measurevar="MElightcyan", groupvars=c("generation","line"))

MElightcyan_summary <- MElightcyan_summary %>%
  mutate(line = case_when(
    line == "treatment" ~ "22 at G1",
    line == "control" ~ "Ambient",
    T ~ line
  ))


#Plot a line interaction plot of a particular module eigengene
ggplot(MElightcyan_summary, aes(x=as.factor(generation), y=MElightcyan, color=line, fill = line, shape = line)) +
  geom_point(size=3, stroke = 1.5 ) +
  geom_errorbar(aes(ymin=MElightcyan-se, ymax=MElightcyan+se), width=.1, alpha = 0.3) +
  geom_line(aes(color=line, group=line, linetype = line)) +
  scale_color_manual(values = c('blue', "red")) +
  scale_shape_manual(values=c(21,23), #labels = c("Ambient","22 at G1")
                     )
                     +
  scale_fill_manual(values=c('blue',"red"), labels = c("Ambient","22 at G1")) +
  xlab("Generation") +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1.5))+
  theme(text = element_text(size = 15)) +
  theme(panel.grid.minor.y = element_blank(), legend.position = "bottom", plot.margin = margin(0,6,0,6))

```

## now the de novo work
#####need to make sure that this runs with only this document loaded
```{r}
# isolate our gene IDs from our eigengene
geneIDs_lightcyan <- module.gene.mapping %>% filter(`bwnet$colors` == 'lightcyan') %>% rownames()

###extract pvalues for each of our generations
interestgenes.and.p <- function(geneIDs, DEseqdata, generation = "G1"){
  CvT <- results(DEseqdata, contrast = c("group", 
                                         sprintf("control%s", generation), 
                                         sprintf("treatment%s", generation)))
  CvT <- CvT[rownames(CvT) %in% geneIDs,]
  CvT <- CvT[order(CvT$padj),] # sort
  CvT <- CvT[!is.na(CvT$padj),] # drop NA
  return(CvT)
}

resG1_CvT_HW <- interestgenes.and.p(geneIDs_lightcyan, dds, 
                                              generation = "G1")
resG2_CvT_HW <- interestgenes.and.p(geneIDs_lightcyan, dds,
                                              generation = "G2")
resG3_CvT_HW <- interestgenes.and.p(geneIDs_lightcyan, dds,
                                              generation = "G3")
resG4_CvT_HW <- interestgenes.and.p(geneIDs_lightcyan, dds,
                                              generation = "G4")
```


```{r}
# function below plots adjusted pvalue, -log(10) and wald statistic
# probably wont use this in final stuff
plotstuff <- function(data, generation){
  res_df <- as.data.frame(data)
  res_df <- res_df[!is.na(res_df$padj), ]
  res_df$negLog10Padj <- -log10(res_df$padj)
  
  # Panel 1: padj distribution
  p1 <- ggplot(res_df, aes(x = padj)) +
  geom_histogram(bins = 50, fill = "skyblue3", color = "black") +
  theme_minimal(base_size = 16) +
  ggtitle(sprintf("Adjusted p-values (padj) Generation %s", generation)) +
  xlab("padj") +
  ylab("Count")

  # Panel 2: -log10(padj)
  p2 <- ggplot(res_df, aes(x = negLog10Padj)) +
    geom_histogram(bins = 50, fill = "tomato3", color = "black") +
    theme_minimal(base_size = 16) +
    ggtitle(sprintf("-log10(padj) Generation %s", generation)) +
    xlab("-log10(padj)") +
    ylab("Count")

# Panel 3: Wald Statistic
  p3 <- ggplot(res_df, aes(x = stat)) +
    geom_histogram(bins = 50, fill = "seagreen3", color = "black") +
    theme_minimal(base_size = 16) +
    ggtitle(sprintf("Wald Statisticn Generation %s", generation)) +
    xlab("stat") +
    ylab("Count")

# Combine vertically with ggpubr
  combined_plot <- ggarrange(
    p1, p2, p3,
    ncol = 1, nrow = 3,
    labels = c("A", "B", "C"),  # optional panel labels
    align = "v",                # vertical alignment
   common.legend = FALSE
  )

# Show combined plot
  return(combined_plot)
}
# this one too
plotstuff2 <- function(data, generation){
  res_df <- as.data.frame(data)
  res_df <- res_df[!is.na(res_df$padj), ]
  res_df$negLog10Padj <- -log10(res_df$padj)
  # Panel D: Wald statistic vs -log10(padj) scatter 
p4 <- ggplot(res_df, aes(x = pvalue, y = negLog10Padj)) +
  geom_point(alpha = 0.4, color = "yellow4") + ylim(c(0,5)) +
  theme_minimal(base_size = 8) +
  ggtitle(sprintf("P-value vs -log10(padj) Generation %s", generation)) +
  xlab("P-value") +
  ylab("-log10(padj)") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "black")

# Panel E: Wald statistic vs -log10(padj) scatter 
p5 <- ggplot(res_df, aes(x = stat, y = negLog10Padj)) +
  geom_point(alpha = 0.4, color = "purple4") + ylim(c(0,5)) +
  theme_minimal(base_size = 8) +
  ggtitle(sprintf("Wald Statistic vs -log10(padj) Generation %s", generation)) +
  xlab("Wald statistic") +
  ylab("-log10(padj)") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "black")


combined_plot <- ggarrange(
  p4, p5,
  ncol = 2, nrow = 1,
  align = "h"
)

# Show combined plot
return(combined_plot)
}

G1_plotstuff <- plotstuff(resG1_CvT_HW, "1")
G2_plotstuff <- plotstuff(resG2_CvT_HW, "2")
G3_plotstuff <- plotstuff(resG3_CvT_HW, "3")
G4_plotstuff <- plotstuff(resG4_CvT_HW, "4")

G1_plotstuff2 <- plotstuff2(resG1_CvT_HW,"1")
G2_plotstuff2 <- plotstuff2(resG2_CvT_HW,"2")
G3_plotstuff2 <- plotstuff2(resG3_CvT_HW,"3")
G4_plotstuff2 <- plotstuff2(resG4_CvT_HW,"4")
```

# TopGO 
```{r}
library(topGO)
library(GO.db)


# Convert to data frame
res_G2_df <- as.data.frame(resG2_CvT_HW)

# Add geneID column from rownames
res_G2_df$geneID <- rownames(res_G2_df)
res_G2_df <- res_G2_df[, c("geneID", setdiff(names(res_G2_df), "geneID"))]

# Read GO terms file (tonsa_go) provided in 
tonsa_go <- read.delim("Genes_GO_terms_output.tsv", header = TRUE) %>%
  slice(-c(1,2)) %>%  # remove header rows if present
  na.omit()  # remove missing GO annotations

# Identify DEGs that have GO annotations
res_G2_GO <- res_G2_df[res_G2_df$geneID %in% tonsa_go$geneID, ]

# Keep only significant DEGs 
res_G2_GO_sig <- subset(res_G2_GO, padj < 0.9 & abs(log2FoldChange) > 0.2)


# Merge with GO annotations
tonsa_stat <- merge(res_G2_GO_sig, tonsa_go, by = "geneID")

# Create gene2GO list which is used in the topgo function 
gene2GO <- tonsa_go %>%
  filter(!is.na(GO) & GO != "") %>%
  group_by(geneID) %>%
  summarise(GO = list(unique(GO)), .groups = "drop")

gene2GO <- setNames(gene2GO$GO, as.character(gene2GO$geneID))

# Split semicolon-separated GO terms
gene2GO <- lapply(gene2GO, function(x) unlist(strsplit(x, ";")))

# Create geneList for TopGO: 1 = DEG, 0 = not DEG
all_genes <- names(gene2GO)
geneList <- setNames(as.integer(all_genes %in% tonsa_stat$geneID), all_genes)

#Build TopGO pipeline 
GOdata <- new("topGOdata",
              description = "G2 DEGs GO enrichment",
              ontology = "BP",
              allGenes = geneList,
              geneSel = function(x) x == 1,
              nodeSize = 5,
              annot = annFUN.gene2GO,
              gene2GO = gene2GO)

#test for enrichment 
result <- runTest(GOdata, algorithm = "parentChild", statistic = "fisher")

# Convert results to table
res_table <- GenTable(GOdata, classicFisher = result,
                      topNodes = length(usedGO(GOdata))) %>%
  mutate(classicFisher = as.numeric(classicFisher),
         Annotated = as.numeric(Annotated),
         Significant = as.numeric(Significant))

topGO_F <- function(data){
  res_G2_df <- as.data.frame(data)
  res_G2_df$geneID <- rownames(res_G2_df)
  res_G2_df <- res_G2_df[, c("geneID", setdiff(names(res_G2_df), "geneID"))]
  tonsa_go <- read.delim("Genes_GO_terms_output.tsv", header = TRUE) %>%
    slice(-c(1,2)) %>%  # remove header rows if present
    na.omit()  # remove missing GO annotations
  res_G2_GO <- res_G2_df[res_G2_df$geneID %in% tonsa_go$geneID, ]
  res_G2_GO_sig <- subset(res_G2_GO, padj < 0.05 & abs(log2FoldChange) > 1)
  tonsa_stat <- merge(res_G2_GO_sig, tonsa_go, by = "geneID")
  gene2GO <- tonsa_go %>%
    filter(!is.na(GO) & GO != "") %>%
    group_by(geneID) %>%
    summarise(GO = list(unique(GO)), .groups = "drop")
  gene2GO <- setNames(gene2GO$GO, as.character(gene2GO$geneID))
  gene2GO <- lapply(gene2GO, function(x) unlist(strsplit(x, ";")))
  all_genes <- names(gene2GO)
  geneList <- setNames(as.integer(all_genes %in% tonsa_stat$geneID), all_genes)
  GOdata <- new("topGOdata",
              description = "G2 DEGs GO enrichment",
              ontology = "BP",
              allGenes = geneList,
              geneSel = function(x) x == 1,
              nodeSize = 5,
              annot = annFUN.gene2GO,
              gene2GO = gene2GO)
  result <- runTest(GOdata, algorithm = "parentChild", statistic = "fisher")
  res_table <- GenTable(GOdata, classicFisher = result,
                      topNodes = length(usedGO(GOdata))) %>%
  mutate(classicFisher = as.numeric(classicFisher),
         Annotated = as.numeric(Annotated),
         Significant = as.numeric(Significant))

  return(res_table)
  
}


```

