---
title: "DEseqOnSpruce"
output: html_document
date: "2025-11-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 
                       "~/projects/eco_genomics_2025/finalproject/")
```

### libraries and data

```{r}
library(DESeq2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(ggpubr)
library(vsn)  
library("pheatmap")
library("vsn")

# import data
setwd("~/projects/eco_genomics_2025")
countsTable <- read.csv("finalproject/mydata/counts_matrix.csv", 
                        header=TRUE, row.names=1)
countsTableRound  <- round(countsTable)
samples <- colnames(countsTableRound)
keep <- !grepl("BLANK", samples)
countsTableRound <- countsTableRound[, keep]
samples <- colnames(countsTableRound)

createMeta <- function(samples){
  parts <- unlist(strsplit(samples, "_"))
  sourcepop <- paste(parts[1], parts[2], sep = "_")
  treatment <- parts[parts %in% c("C", "H", "D")]
  treatmentmap <- c(
    C = "Control",
    H = "Heat",
    D = "Drought"
  )
  treatments <- unname(treatmentmap[treatment])
  treatment_string <- paste(unique(treatments), collapse = "; ")
  lasttreatindex <- max(which(
    parts %in% c("C", "H", "D")))
  day <- as.integer(parts[lasttreatindex + 1])
  data.frame(source_population = sourcepop,
             treatment = treatment_string,
             day_harvest = day,
             stringsAsFactors = FALSE)
}
meta_list <- lapply(samples, createMeta)
metadataSpruceRNA <- bind_rows(meta_list) 
metadataSpruceRNA$samples <- samples
rownames(metadataSpruceRNA) <- metadataSpruceRNA$samples
metadataSpruceRNA <- metadataSpruceRNA %>%
  mutate(source_climate = case_when(
    source_population %in% c("ASC_06","BRU_05",
                             "ESC_01", "XBM_07",
                             "NOR_02") ~ "HotDry",
    source_population %in% c("CAM_02", "JAY_02",
                             "KAN_04", "LOL_02",
                             "MMF_13") ~ "CoolWet"
  ))

```

### data distribution

```{r}
# Let's see how many reads we have from each sample
colSums(countsTableRound)
mean(colSums(countsTableRound))

barplot(colSums(countsTableRound), names.arg=colnames(countsTableRound),cex.names=0.5, las=3,ylim=c(0,2500000)) #change dims
abline(h=mean(colSums(countsTableRound)), col="blue", lwd=2)

# the average number of counts per gene
rowSums(countsTableRound)
mean(rowSums(countsTableRound)) # 
median(rowSums(countsTableRound)) #

apply(countsTableRound,2,mean) # 2 in the apply function does the action across columns
apply(countsTableRound,1,mean) # 1 in the apply function does the action across rows
hist(apply(countsTableRound,1,mean),xlim=c(0,14), ylim=c(0,60000),breaks=1000000)
```

### DEseq2

```{r}
dds_sourceVday <- DESeqDataSetFromMatrix(countData = countsTableRound,
                              colData=metadataSpruceRNA, 
                              design= ~ source_climate + day_harvest)
dim(dds_sourceVday)
dds_sourceVday <- dds_sourceVday[rowSums(counts(dds_sourceVday) >= 15) >= 18,]
dds_sourceVday <- DESeq(dds_sourceVday)
resultsNames(dds_sourceVday)
# source climate versus treatment
dds_sourceVtreatment <- DESeqDataSetFromMatrix(countData = countsTableRound,
                              colData=metadataSpruceRNA, 
                              design= ~ source_climate + treatment)
dim(dds_sourceVtreatment)
dds_sourceVtreatment <- dds_sourceVtreatment[rowSums(counts(dds_sourceVtreatment) >= 15) >= 18,]
dds_sourceVtreatment <- DESeq(dds_sourceVtreatment)
resultsNames(dds_sourceVtreatment)
```

### sample clustering and visualization

```{r}
# this gives log2(n + 1)
ntd <- normTransform(dds_sourceVday)
meanSdPlot(assay(ntd))

# Variance stabilizing transformation
vsd <- vst(dds_sourceVday, blind=FALSE)
meanSdPlot(assay(vsd))
```

### heatmap

```{r}
sampleDists <- dist(t(assay(vsd)))

library("RColorBrewer")
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$line, vsd$generation, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Purples")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```


### cluster tree

```{r}
sampleTree <- hclust(dist(sampleDists), method="average")
# plot
plot(sampleTree, main="Sample clustering to detect outliers", sub="", xlab="",cex.lab=1.5, cex.axis=1.5, cex.main=2)
```

### PCA

```{r}
# first transform the data for plotting using variance stabilization
vsd <- vst(dds_sourceVday, blind=FALSE)

pcaData <- plotPCA(vsd,
                   intgroup=c("source_climate","day_harvest"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData,"percentVar"))

ggplot(pcaData, aes(PC1, PC2, color=day_harvest,
                    shape=source_climate)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() +
  theme_linedraw()
```

### PCA2

```{r}
ggplot(pcaData, aes(PC1, PC2)) +
  geom_point(size=5, stroke = 1, aes(fill=source_climate, shape=treatment, alpha = day_harvest)) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) +
  #scale_shape_manual(values=c(21,22,23,24), labels = c("G1", "G2","G3", "G4"))+
  #scale_shape_manual(values=c(21,22,23,24) -> circle, square, diamond, triangle
  scale_fill_manual(values=c('#6699CC',"#CC3333"), labels = c("Control", "Treatment"))+
  #scale_alpha_manual(values = c(1.0, 0.8, 0.7, 0.3), guide = "none") +
  theme_linedraw() +
  theme(legend.position = "none") +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 3))+
  theme(text = element_text(size = 20)) +
  theme(legend.title = element_blank()) +
  theme_linedraw()
```
  - there doesn't look to be much clustering in the PCA based on any of the variables 
  - lets try isolating specific contrasts
  
### source_climate versus treatment
```{r}
metadataSpruceRNA$day_harvest <- as.character(metadataSpruceRNA$day_harvest)
dds_sourceVday$source_climate <- relevel(dds_sourceVday$source_climate, ref = "HotDry")
dds_sourceVtreatment$treatment <- relevel(dds_sourceVtreatment$treatment, ref = "Control")
#dds_sourceVday$day_harvest <- relevel(dds_sourceVday$day_harvest, ref = "0")


dds_sourceVday$group <- factor(paste0(dds_sourceVday$source_climate, dds_sourceVday$day_harvest))
design(dds_sourceVday) <- ~ group
dds_sourceVday <- DESeq(dds_sourceVday)
resultsNames(dds_sourceVday)
# source V treatment
dds_sourceVtreatment$group <- factor(paste0(dds_sourceVday$source_climate, dds_sourceVday$treatment))
design(dds_sourceVtreatment) <- ~ group
dds_sourceVtreatment <- DESeq(dds_sourceVtreatment)
resultsNames(dds_sourceVtreatment)

#source climates compared, at all time points
coolwetVhotdry_t0 <- results(dds_sourceVday,
                             contrast=c("group","CoolWet0","HotDry0"), 
                             alpha = 0.05)
coolwetVhotdry_t5 <- results(dds_sourceVday,
                             contrast=c("group","CoolWet5","HotDry5"), 
                             alpha = 0.05)
coolwetVhotdry_t10 <- results(dds_sourceVday,
                              contrast=c("group","CoolWet10","HotDry10"), 
                              alpha = 0.05)
#temporal
hotdry_0v5 <- results(dds_sourceVday, contrast=c("group","HotDry0","HotDry5"), alpha = 0.05)
hotdry_0v10 <- results(dds_sourceVday, contrast=c("group","HotDry0","HotDry10"), alpha = 0.05)
hotdry_5v10 <- results(dds_sourceVday, contrast=c("group","HotDry5","HotDry10"), alpha = 0.05)
coolwet_0v5 <- results(dds_sourceVday, contrast=c("group","CoolWet0","CoolWet5"), alpha = 0.05)
coolwet_0v10 <- results(dds_sourceVday, contrast=c("group","CoolWet0","CoolWet10"), alpha = 0.05)
coolwet_5v10 <- results(dds_sourceVday, contrast=c("group","CoolWet5","CoolWet10"), alpha = 0.05)

#treatments
hotdry_CvH <- results(dds_sourceVtreatment,
                      contrast=c("group","HotDryControl","HotDryHeat"), 
                      alpha = 0.05)
hotdry_CvDH <- results(dds_sourceVtreatment,
                      contrast=c("group","HotDryControl","HotDryDrought; Heat"), 
                      alpha = 0.05)
hotdry_HvDH <- results(dds_sourceVtreatment,
                      contrast=c("group","HotDryHeat","HotDryDrought; Heat"), 
                      alpha = 0.05)
coolwet_CvH <- results(dds_sourceVtreatment,
                      contrast=c("group","CoolWetControl","CoolWetHeat"), 
                      alpha = 0.05)
coolwet_CvDH <- results(dds_sourceVtreatment,
                      contrast=c("group","CoolWetControl","CoolWetDrought; Heat"), 
                      alpha = 0.05)
coolwet_HvDH <- results(dds_sourceVtreatment,
                      contrast=c("group","CoolWetHeat","CoolWetDrought; Heat"), 
                      alpha = 0.05)
```

### plot above results

```{r}
plotMA(coolwetVhotdry_t0, ylim=c(-5,5))
plotMA(coolwetVhotdry_t5, ylim=c(-5,5))
plotMA(coolwetVhotdry_t10, ylim=c(-5,5))
plotMA(hotdry_0v5, ylim=c(-5,5))
plotMA(hotdry_0v10, ylim=c(-5,5))
plotMA(hotdry_5v10, ylim=c(-5,5))
plotMA(coolwet_0v5, ylim=c(-5,5))
plotMA(coolwet_0v10, ylim=c(-5,5))
plotMA(coolwet_5v10, ylim=c(-5,5))
plotMA(hotdry_CvH, ylim=c(-5,5))
plotMA(hotdry_CvDH, ylim=c(-5,5))
plotMA(hotdry_HvDH, ylim=c(-5,5))
plotMA(coolwet_CvH, ylim=c(-5,5))
plotMA(coolwet_CvDH, ylim=c(-5,5))
plotMA(coolwet_HvDH, ylim=c(-5,5))
```

### volcano plots
```{r}
# Make a dataframe
vol <- function(data, title){
res_df <- as.data.frame(data)
res_df <- res_df[!is.na(res_df$padj), ]
res_df$significance <- "Not Sig"
res_df$significance[res_df$padj < 0.05 & res_df$log2FoldChange > 0] <- "Up"
res_df$significance[res_df$padj < 0.05 & res_df$log2FoldChange < 0] <- "Down"
ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
  geom_point(alpha = 0.7, size = 1) +
  scale_color_manual(values = c("Up" = "red", "Down" = "blue", "Not Sig" = "grey")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  labs(
    title = title,
    x = expression(log[2]~Fold~Change),
    y = expression(-log[10]~Adjusted~p~value)
  ) +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = "none",
        plot.title = element_text(hjust = 0.5),
        axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 12))
}

vol(coolwetVhotdry_t0, "CWvHT t0")
vol(coolwetVhotdry_t5, "CWvHT t5")
vol(coolwetVhotdry_t10, "CWvHT t10")


HD_0v5 <- vol(hotdry_0v5, "t0 vs t5") +
  scale_y_continuous(breaks = c(0,3,6,9))
HD_0v10 <- vol(hotdry_0v10, "t0 vs t10") + theme(axis.title.y = element_blank())
HD_5v10 <- vol(hotdry_5v10, "t5 vs t10") + theme(axis.title.y = element_blank())
CW_0v5 <- vol(coolwet_0v5, "t0 vs t5")
CW_0v10 <- vol(coolwet_0v10, "t0 vs t10")+ theme(axis.title.y = element_blank())
CW_5v10 <- vol(coolwet_5v10, "t5 vs t10")+ theme(axis.title.y = element_blank())


vol(hotdry_CvH, "CvH HD")
vol(hotdry_CvDH, "Control versus Drought & Heat treatments (source = hot/dry)")
vol(hotdry_HvDH, "HvDH HD")
vol(coolwet_CvH, "CvH CW")
vol(coolwet_CvDH, "Control versus Drought & Heat treatments (source = cool/wet)")
vol(coolwet_HvDH, "HvDH CW")

volcaner <- ggarrange(HD_0v5, HD_0v10, HD_5v10,
          CW_0v5, CW_0v10, CW_5v10,
          widths = c(1.1,1,1),
          common.legend = T, 
          legend = "bottom")
ggsave("volcano.tif", plot = volcaner)
```

## individual genes
### for cool 
```{r}
heat_map <- function(data, dds, numgenes, 
                     factor1, factor2){
topgenes <- head(rownames(data),numgenes)
mat <- assay(vsd)[topgenes,]
mat <- mat - rowMeans(mat)
df <- as.data.frame(colData(dds)[,c(factor1,factor2)])

#pheatmap(mat, annotation_col=df)
pheatmap(mat, annotation_col=df, cluster_cols = F,
         show_colnames = F)
}
# above plot it pretty good,could be a figure
heat_sourceVday <- heat_map(hotdry_0v10,
                            dds_sourceVtreatment, 10,
                            "treatment",
                            "day_harvest")

```


## heat map of LFC of gene across cool wet and day harvested
```{r}
sig_genes <- rownames(coolwet_0v10[which(coolwet_0v10$padj < 0.05 & !is.na(coolwet_0v10$padj)), ])

# Combine into a data frame
lfc_mat <- data.frame(
  "0v10" = coolwet_0v10[sig_genes, "log2FoldChange"],
  "0v5" = coolwet_0v5[sig_genes, "log2FoldChange"],
  "5v10" = coolwet_5v10[sig_genes, "log2FoldChange"]
)

# Remove any rows with missing values (some genes may not be in all result tables)
lfc_mat <- na.omit(lfc_mat)

pheatmap(lfc_mat,
         scale = "row",             # scales each gene’s LFCs across generations
         color = colorRampPalette(c("blue", "white", "red"))(100),
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         show_rownames = FALSE,
         main = "LFC (cool/wet across day harvested)")
```

## same but hot dry
```{r}
sig_genes_hotdrytemporal <- rownames(hotdry_0v10[which(hotdry_0v10$padj < 0.05 & !is.na(hotdry_0v10$padj)), ])

# Combine into a data frame
lfc_mat_hotdry <- data.frame(
  "0v10" = hotdry_0v10[sig_genes_hotdrytemporal, "log2FoldChange"],
  "0v5" = hotdry_0v5[sig_genes_hotdrytemporal, "log2FoldChange"],
  "5v10" = hotdry_5v10[sig_genes_hotdrytemporal, "log2FoldChange"]
)

# Remove any rows with missing values (some genes may not be in all result tables)
lfc_mat_hotdry <- na.omit(lfc_mat_hotdry)

pheatmap(lfc_mat_hotdry,
         scale = "row",             # scales each gene’s LFCs across generations
         color = colorRampPalette(c("blue", "white", "red"))(100),
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         show_rownames = FALSE,
         main = "LFC (hot/dry across day harvested)")
```

## cool wet v hot dry treatments

```{r}
sig_genes_treatmenthotdry <- rownames(hotdry_CvDH[which(hotdry_CvDH$padj < 0.05 & !is.na(hotdry_CvDH$padj)), ])

# Combine into a data frame
lfc_mat_treatmenthotdry <- data.frame(
  CvH = hotdry_CvH[sig_genes_treatmenthotdry, "log2FoldChange"],
  CvDH = hotdry_CvDH[sig_genes_treatmenthotdry, "log2FoldChange"],
  HvDH = hotdry_HvDH[sig_genes_treatmenthotdry, "log2FoldChange"]
)

# Remove any rows with missing values (some genes may not be in all result tables)
lfc_mat_treatmenthotdry <- na.omit(lfc_mat_treatmenthotdry)

pheatmap(lfc_mat_treatmenthotdry,
         scale = "row",             # scales each gene’s LFCs across generations
         color = colorRampPalette(c("blue", "white", "red"))(100),
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         show_rownames = FALSE,
         main = "LFC (hot/dry across treatment)")
```


# coolwet

```{r}
sig_genes_treatmentcoolwet <- rownames(coolwet_CvDH[which(coolwet_CvDH$padj < 0.05 & !is.na(coolwet_CvDH$padj)), ])

# Combine into a data frame
lfc_mat_treatmentcoolwet <- data.frame(
  CvH = coolwet_CvH[sig_genes_treatmentcoolwet, "log2FoldChange"],
  CvDH = coolwet_CvDH[sig_genes_treatmentcoolwet, "log2FoldChange"],
  HvDH = coolwet_HvDH[sig_genes_treatmentcoolwet, "log2FoldChange"]
)

# Remove any rows with missing values (some genes may not be in all result tables)
lfc_mat_treatmentcoolwet <- na.omit(lfc_mat_treatmentcoolwet)

pheatmap(lfc_mat_treatmentcoolwet,
         scale = "row",             # scales each gene’s LFCs across generations
         color = colorRampPalette(c("blue", "white", "red"))(100),
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         show_rownames = FALSE,
         main = "LFC (cool/wet across treatment)")
```



## getting differentially expressed genes

```{r}
coolwet_0v10$padj[is.na(coolwet_0v10$padj)] <- 0.5
degs_coolwet_0v10 <- row.names(coolwet_0v10[coolwet_0v10$padj < 0.05,])

hotdry_0v10$padj[is.na(hotdry_0v10$padj)] <- 0.5
degs_hotdry_0v10 <- row.names(hotdry_0v10[hotdry_0v10$padj < 0.05,])

coolwet_0v5$padj[is.na(coolwet_0v5$padj)] <- 0.5
degs_coolwet_0v5 <- row.names(coolwet_0v5[coolwet_0v5$padj < 0.05,])

hotdry_0v5$padj[is.na(hotdry_0v5$padj)] <- 0.5
degs_hotdry_0v5 <- row.names(hotdry_0v5[hotdry_0v5$padj < 0.05,])


commondegs <- degs_coolwet_0v10[degs_coolwet_0v10 %in% degs_hotdry_0v10]
filtered_degs_hotdry10 <- hotdry_0v10[rownames(hotdry_0v10) %in% commondegs,]
filtered_degs_coolwet10 <- coolwet_0v10[rownames(coolwet_0v10) %in% commondegs,]


## treatment between source climates
coolwet_CvDH$padj[is.na(coolwet_CvDH$padj)] <- 0.5
degs_coolwet_CvDH <- row.names(coolwet_CvDH[coolwet_CvDH$padj < 0.05,])

hotdry_CvDH$padj[is.na(hotdry_CvDH$padj)] <- 0.5
degs_hotdry_CvDH <- row.names(hotdry_CvDH[hotdry_CvDH$padj < 0.05,])

commonTreatmentDegs <- degs_hotdry_CvDH[degs_hotdry_CvDH %in% degs_coolwet_CvDH]
filtered_degs_coolwetDH <- coolwet_CvDH[
  rownames(coolwet_CvDH) %in% commonTreatmentDegs,]
filtered_degs_hotdryDH <- hotdry_CvDH[
  rownames(hotdry_CvDH) %in% commonTreatmentDegs,]

dfhotdryTreatment <- data.frame(
  row.names = c(rownames(filtered_degs_hotdryDH)), 
  LFC = filtered_degs_hotdryDH$log2FoldChange
)
dfcoolwetTreatment <- data.frame(
  row.names = c(rownames(filtered_degs_coolwetDH)),
  LFC = filtered_degs_coolwetDH$log2FoldChange
)

TreatmentPlotDF <- merge(dfcoolwetTreatment, dfhotdryTreatment, 
                         by = "row.names", all = T)
TreatmentPlotDF <- TreatmentPlotDF %>%
  rename(hotdry = LFC.y,
         coolwet = LFC.x)
```

- steve and I spoke about a plot that could be good
 - taking log fold change for significant genes among a subset of a treatment
  - ex: similar degs for cool/wet heat at day 10 V hot/dry heat day 10
  
- result: when comparing coolwet v hot try at day 10, there are 5 genes that are common :(
 - at day 5, there are only 4 
  
```{r}
library(dplyr)
dfhotdry <- data.frame(
  row.names = c(rownames(filtered_degs_hotdry10)), 
  LFC = filtered_degs_hotdry10$log2FoldChange
)
dfcoolwet <- data.frame(
  row.names = c(rownames(filtered_degs_coolwet10)),
  LFC = filtered_degs_coolwet10$log2FoldChange
)
plotdf <- merge(dfcoolwet, dfhotdry, by = "row.names", all = T)

plotdf <- plotdf %>%
  rename(hotdry = LFC.x) 
plotdf <- plotdf %>%
  rename(coolwet = LFC.y)
```

### LFC temporal source climate
```{r}
temporalPlot <- ggplot() +
  geom_point(data = plotdf, aes(x = hotdry, y = coolwet),
             size = 3, color = "green4") +
  geom_smooth(data = plotdf, aes(x = hotdry, y = coolwet),
              method = "lm", formula = y~x +0, fullrange = T, level = 0.3,
              color = "green4") +
  geom_abline(intercept = 0, slope = 1, alpha = 0.3) +
  geom_abline(intercept = 0, slope = 0, alpha = 0.5) +
  geom_vline(xintercept = 0, alpha = 0.5) +
  geom_density(data = hotdry_0v10[hotdry_0v10$padj < 0.1,], 
               aes(x = log2FoldChange, y = after_stat(ndensity)),
                   fill = "tan",
               position = position_nudge(y = -3.76),
               bw = 0.1) +
  geom_density(data = coolwet_0v10[coolwet_0v10$padj < 0.1,], 
               aes(y = log2FoldChange, x = after_stat(ndensity)),
                   fill = "lightblue",
               position = position_nudge(x = -3.28),
               bw = 0.1) +
  
  geom_density(data = hotdry_0v10[hotdry_0v10$padj < 0.1,], 
               aes(y = log2FoldChange, x = after_stat(ndensity)),
                   fill = "tan",
               position = position_nudge(x = -3.28),
               bw = 0.1, alpha = 0.1, linewidth = 0.1) +
  geom_density(data = coolwet_0v10[coolwet_0v10$padj < 0.1,], 
               aes(x = log2FoldChange, y = after_stat(ndensity)),
                   fill = "lightblue",
               position = position_nudge(y = -3.76),
               bw = 0.1, alpha = 0.1, linewidth = 0.1) +
  coord_cartesian(xlim = c(-3,3), ylim = c(-3.5,2)) +
  theme(panel.border = element_rect(color = "black")) +
  labs(x = "LFC day 0 vs day 10 (Hot/Dry Source)",
       y = "LFC day 0 vs day 10 (Cool/wet Source)",
       title = "LFC of Common Temporal DEGs against Source Climate")
```
### LFC treatment and source climate
```{r}
ggplot() +
  geom_point(data = TreatmentPlotDF, aes(x = hotdry, y = coolwet),
             size = 3, color = "green4") +
  geom_smooth(data = TreatmentPlotDF, aes(x = hotdry, y = coolwet),
              method = "lm", formula = y~x +0, fullrange = T, level = 0.5,
              xseq = -4:0, color = "green4") +
  #volcaner
  geom_point(data = hotdry_CvDH, aes(x = log2FoldChange, y = -log10(padj)-3.7), alpha = 0.2, color = "tan") +
  geom_point(data = coolwet_CvDH, aes(y = log2FoldChange, x = -log10(padj)-3.2), alpha = 0.2, color = "lightblue") +
  geom_abline(intercept = 0, slope = 1, alpha = 0.3) +
  geom_abline(intercept = 0, slope = 0, alpha = 0.5) +
  geom_vline(xintercept = 0, alpha = 0.5) +
  geom_density(data = hotdry_CvDH[hotdry_CvDH$padj < 0.1,], 
               aes(x = log2FoldChange, y = after_stat(ndensity)),
                   fill = "tan",
               position = position_nudge(y = -3.75),
               bw = 0.1) +
  geom_density(data = coolwet_CvDH[coolwet_CvDH$padj < 0.1,], 
               aes(y = log2FoldChange, x = after_stat(ndensity)),
                   fill = "lightblue",
               position = position_nudge(x = -3.251),
               bw = 0.1) +
  #overlays
  geom_density(data = hotdry_CvDH[hotdry_CvDH$padj < 0.1,], 
               aes(y = log2FoldChange, x = after_stat(ndensity)),
                   fill = "tan",
               position = position_nudge(x = -3.251),
               bw = 0.1, alpha = 0.1, linewidth = 0.1) +
  geom_density(data = coolwet_CvDH[coolwet_CvDH$padj < 0.1,], 
               aes(x = log2FoldChange, y = after_stat(ndensity)),
                   fill = "lightblue",
               position = position_nudge(y = -3.75),
               bw = 0.1, alpha = 0.1, linewidth = 0.1) +
  coord_cartesian(xlim = c(-3,2), ylim = c(-3.5,2)) +
  theme(panel.border = element_rect(color = "black")) +
  labs(x = "LFC Control vs Drought + Heat (Hot/Dry Source)",
       y = "LFC Control vs Drought + Heat (Cool/wet Source)",
       title = "LFC of Common Treatment DEGs against Source Climate")
```


```{r}
ggsave("TemporalPlot.tif", plot = temporalPlot)
ggsave("TreatmentPlot.tif", plot = treatmentPlot)
```


```{r}
ggplot() +
  geom_point(data = TreatmentPlotDF, aes(x = hotdry, y = coolwet),
             size = 3, color = "green4") +
  geom_smooth(data = TreatmentPlotDF, aes(x = hotdry, y = coolwet),
              method = "lm", formula = y~x +0, fullrange = T, level = 0.5,
              xseq = -4:0, color = "green4") +
  geom_abline(intercept = 0, slope = 1, alpha = 0.3) +
  geom_abline(intercept = 0, slope = 0, alpha = 0.5) +
  geom_vline(xintercept = 0, alpha = 0.5) +
  geom_point(data = hotdry_CvDH, aes(x = log2FoldChange, y = -log10(padj)-3.7), alpha = 0.2, color = "tan") +
  coord_cartesian(xlim = c(-3,2), ylim = c(-3.5,2)) +
  theme(panel.border = element_rect(color = "black")) +
  labs(x = "LFC Control vs Drought + Heat (Hot/Dry Source)",
       y = "LFC Control vs Drought + Heat (Cool/wet Source)",
       title = "LFC of Common Treatment DEGs against Source Climate")
```



Gene ontology

```{r}
library(stringr)
z <- read.table("/gpfs1/cl/ecogen/pbio6800/PopulationGenomics/ref_genome/Pmariana/annotation/GO_annotations/Pmariana_GO_annotations_sorted_join.txt", header=F, fill=T)

z_new <- z %>%
  mutate(V7=ifelse(V7 %in% "", V2, V7))
colnames(z_new) <- c("gene", "start", "stop", "geneID", 
                     "strand", "parent", "marker")

x <- read.table("/gpfs1/cl/ecogen/pbio6800/GroupProjects/hotConifers/RS_bt2", 
                header=F)
x2 <- x[, c("V1", "V4")]
colnames(x2) <- c("marker", "geneID")

######
GO_table <- read.delim("/gpfs1/cl/ecogen/pbio6800/GroupProjects/hotConifers/GO_table_spruce.tsv",
                       header = T, sep = "\t")
GO_table <- GO_table %>%
  distinct(marker, .keep_all = TRUE) #remove duplicates

GO_table_try <- merge(x2, GO_table, by ="marker") #add geneIDs to the GO terms

zGO <- GO_table_try[, c("geneID.x", "gene")] #filter out nonuseful columns

colnames(zGO) <- c("geneID", "GO") #rename column names

zGO <- zGO %>%
  mutate(geneID = str_remove(geneID, "ID="))

gene2GO_spruce <- zGO %>%
  filter(!is.na(GO) & GO != "") %>%
  group_by(geneID) %>%
  summarise(GO = list(unique(GO)), .groups = "drop") %>%
  with(setNames(zGO$GO, zGO$geneID))

gene2GO_spruce <- lapply(gene2GO_spruce, function(x) unlist(strsplit(x, ","))) #this is taking quite a while
#this above object should be good for direct input into topGO

```

# sample GO work
```{r}
topGOshbang_spruce <- function(data){
library(topGO)
library(GO.db)

# Convert to data frame
GOtest <- as.data.frame(data)

# Add geneID column from rownames
GOtest$geneID <- rownames(GOtest)
GOtest <- GOtest[, c("geneID", setdiff(names(GOtest), "geneID"))]

# Identify DEGs that have GO annotations
GOtest <- GOtest[GOtest$geneID %in% zGO$geneID, ]

# Keep only significant DEGs 
GOtest_sig <- subset(GOtest, padj < 0.05 & abs(log2FoldChange) > 0.5)


# Merge with GO annotations
spruce_stat <- merge(GOtest_sig, zGO, by = "geneID")

spruce_stat <- spruce_stat %>%
  distinct(geneID, .keep_all = TRUE) #remove duplicates

# Create geneList for TopGO: 1 = DEG, 0 = not DEG
all_genes <- names(gene2GO_spruce)
geneList <- setNames(as.integer(all_genes %in% spruce_stat$geneID), all_genes)

#Build TopGO pipeline 
GOdata <- new("topGOdata",
              description = "GO enrichment",
              ontology = "BP",
              allGenes = geneList,
              geneSel = function(x) x == 1,
              nodeSize = 5,
              annot = annFUN.gene2GO,
              gene2GO = gene2GO_spruce)

#test for enrichment 
result <- runTest(GOdata, algorithm = "parentChild", statistic = "fisher")

# Convert results to table
res_table <- GenTable(GOdata, classicFisher = result,
                      topNodes = length(usedGO(GOdata))) %>%
  mutate(classicFisher = as.numeric(classicFisher),
         Annotated = as.numeric(Annotated),
         Significant = as.numeric(Significant))


#plot!

# Keep GO terms with enough but not too many genes
filtered_GO_results <- res_table %>%
  filter(Annotated >= 5 & Annotated < 500) %>%  # adjust as needed
  group_by(Term) %>%
  slice_min(classicFisher, n = 1) %>%  # keep only the most significant per Term
  ungroup() %>%
  arrange(classicFisher)

return(filtered_GO_results)
}

TopGO_plot <- function(df, title, categories = 30) {
  df %>%
    slice_head(n = categories) %>%
    ggplot(aes(x = -log10(classicFisher),
               y = reorder(Term, -log10(classicFisher)),
               size = Significant,
               color = -log10(classicFisher))) +
    geom_point(alpha = 0.8) +
    scale_color_gradientn(
      colors = c("pink", "orchid", "purple", "purple4"),
      name = "-log10(p-value)"
    ) +
    scale_size_continuous(name = "# Significant Genes") +
    xlab("-log10-classicFisher") +
    ylab("Biological Function") +
    ggtitle(title) +
    theme_minimal() +
    theme(plot.margin = margin(10, 100, 10, 10),
          axis.text.y = element_text(size = 10, hjust = 1)) +
    coord_cartesian(clip = "off")
}

```


```{r}

#hotdry
hotdry_0v10_GO <- topGOshbang_spruce(hotdry_0v10)
hotdry_0v5_GO <- topGOshbang_spruce(hotdry_0v5)
hotdry_5v10_GO <- topGOshbang_spruce(hotdry_5v10)

#coolwet
coolwet_0v10_GO <- topGOshbang_spruce(coolwet_0v10)
coolwet_0v5_GO <- topGOshbang_spruce(coolwet_0v5)
coolwet_5v10_GO <- topGOshbang_spruce(coolwet_5v10)

#hotdry
TopGO_plot(hotdry_0v10_GO, title = "0v10", categories = 10)
TopGO_plot(hotdry_0v5_GO, title = "0v5", categories = 10)
TopGO_plot(hotdry_5v10_GO, title = "5v10", categories = 10)

#coolwet
TopGO_plot(coolwet_0v10_GO, title = "0v10", categories = 10)
TopGO_plot(coolwet_0v5_GO, title = "0v5", categories = 10)
TopGO_plot(coolwet_5v10_GO, title = "5v10", categories = 10)
```


# common GO
```{r}

temporal_common_toGO <- hotdry_0v10[rownames(hotdry_0v10) %in% commondegs,]
treatment_common_toGO <- coolwet_0v10[rownames(coolwet_0v10) %in% commonTreatmentDegs,]

```

#revigo
```{r}
revigo <- function(data, color = "color"){
# Select only the columns REVIGO needs
revigo_input <- data[, c("GO.ID", "classicFisher")]

sigRes <- revigo_input[as.numeric(revigo_input$classicFisher) < 0.2, ]

# Rename columns for clarity
colnames(sigRes) <- c("GO.ID", "pValue")

# Save as txt to upload to REVIGO
write.table(sigRes, file = sprintf("topGOsig_for_REVIGO_%s.txt", color), sep = "\t", row.names = FALSE, quote = FALSE)
}

setwd("~/projects/eco_genomics_2025/finalproject/mydata/")

revigo(hotdry_0v10_GO, color = "HD0v10")
revigo(hotdry_0v5_GO, color = "HD0v5")
revigo(hotdry_5v10_GO, color = "HD5v10")

revigo(coolwet_0v10_GO, color = "CW0v10")
revigo(coolwet_5v10_GO, color = "CW0v5")
revigo(coolwet_5v10_GO, color = "CW5v10")
```
 
#cool revigo plot

```{r}
geom_point_color <- function(color, color_id = "brown") {

  p <- geom_point(data = color, aes(PC_0,PC_1), 
             color = color_id,
             size = color$LogSize * 1.75, 
             alpha = color$Valuescaled)
    
return(p)
  
}

library(readr)

setwd("~/projects/eco_genomics_2025/finalproject/myresults/")

CW05 <- read_tsv("CW05.tsv")
CW010 <- read_tsv("CW010.tsv")
CW510 <- read_tsv("CW510.tsv")

HD05 <- read_tsv("HD05.tsv")
HD010 <- read_tsv("HD010.tsv")
HD510 <- read_tsv("HD510.tsv")

#####
CW05 <- CW05 %>%
  mutate(Valuescaled = Value /min(Value))
CW05 <- CW05 %>%
  filter(PC_0 != "null") %>%
  filter(PC_1 != "null")
CW05$source <- "Cool/Wet 0v5"

CW010 <- CW010 %>%
  mutate(Valuescaled = Value /min(Value))
CW010 <- CW010 %>%
  filter(PC_0 != "null") %>%
  filter(PC_1 != "null")
CW010$source <- "Cool/Wet 0v10"

CW510 <- CW510 %>%
  mutate(Valuescaled = Value /min(Value))
CW510 <- CW510 %>%
  filter(PC_0 != "null") %>%
  filter(PC_1 != "null")
CW510$source <- "Cool/Wet 5v10"

HD010 <- HD010 %>%
  mutate(Valuescaled = Value /min(Value))
HD010 <- HD010 %>%
  filter(PC_0 != "null") %>%
  filter(PC_1 != "null")
HD010$source <- "Hot/Dry 0v10"

HD05 <- HD05 %>%
  mutate(Valuescaled = Value /min(Value))
HD05 <- HD05 %>%
  filter(PC_0 != "null") %>%
  filter(PC_1 != "null")
HD05$source <- "Hot/Dry 0v5"

HD510 <- HD510 %>%
  mutate(Valuescaled = Value /min(Value))
HD510 <- HD510 %>%
  filter(PC_0 != "null") %>%
  filter(PC_1 != "null")
HD510$source <- "Hot/Dry 5v10"
######

totaldf <- rbind(CW010, CW05, CW510,
                 HD010, HD05, HD510)


CW05p <- geom_point_color(CW05, "lightblue1")
CW010p <- geom_point_color(CW010, "blue")
CW510p <- geom_point_color(CW510, "lightblue4")

HD05p <- geom_point_color(HD05, "tan")
HD010p <- geom_point_color(HD010, "red3")
HD510p <- geom_point_color(HD510, "red1")

totaldf$PC_0 <- as.numeric(totaldf$PC_0)
totaldf$PC_1 <- as.numeric(totaldf$PC_1)

ggplot() +
  geom_point(data = totaldf, aes(PC_0, PC_1, 
                                 group = source, 
                                 color = source),
             size = totaldf$LogSize * 1.75, 
             alpha = totaldf$Valuescaled) +
  labs(x = "Semantic Space X", y = "Semantic Space Y") +
  theme() +
  scale_color_manual(values = c("forestgreen", "lightblue", "blue", 
                       "tan", "orange2", "red1"))
  
  
  
  
  
```


